STA302 Fall 2024 Final Project Part 1: Research Proposal and Data Introduction

### 2024-12-05 final_paper/paper.qmd
Save changes to final report files

# Preliminary Results

Problem Definition
Focusing on the *New Housing Price Index* (NHPI) as the response variable, We will analyze economic, construction, and financial predictors such as construction activity, interest rates, and GDP.

## **1. Data Preparation**

- **Tools Used**: Analysis was conducted in **R** using `tidyverse` for data manipulation and `car` for diagnostics.

- **Steps Taken**:  
  - Defined **New Housing Price Index (NHPI)** as the response variable and identified predictors:  
  1. **Detached Absorption (Quarterly Average)**  
  2. **Semi Absorption (Quarterly Average)**  
  3. **Detached Unabsorbed Homes (Quarterly Average)**  
  4. **Semi Unabsorbed Homes (Quarterly Average)**  
  5. **GDP (Quarterly Average)**  
  6. **CPI (Quarterly Average)**  
  7. **Starting Detached Construction**  
  8. **Starting Semi-Detached Construction**  
  9. **Under Construction Detached Homes**  
  10. **Under Construction Semi-Detached Homes**  
  11. **Completed Construction Semi-Detached**  
  12. **Completed Construction Detached**  
  13. **Rate Hikes**
  - Removed missing values and outliers using Cook’s Distance.
  - Merged datasets from 1997–2016, ensuring quarterly granularity.  
  - **Split the Data**:  
    - Divided the dataset into training and test/validation sets.  
    - Used a common split, such as 50/50, to ensure both subsets are sufficiently large.  
    - Performed the split randomly to avoid sampling bias.  
```{r}
# Load necessary libraries
library(tidyverse)
library(here)
library(caret)

@@

## **2. Exploratory Data Analysis (EDA)**
- **Tools Used**: `ggplot2` for visualization and `corrplot` for examining correlations.  
- **Key Insights**:  
  - **Scatterplots** indicated potential linear relationships between predictors and NHPI.  
  - **Variance Inflation Factors (VIFs)** flagged multicollinearity among construction-related predictors, necessitating feature selection or transformation.  
  - Outliers were identified, influencing NHPI trends, and removed to enhance model accuracy.  
  - **Explore Training and Test Sets**:  
    - Performed EDA on both `training_set` and `test_set` to ensure consistency and reliability.  
    - Compared distributions, means, variances, and shapes of variables across both sets to ensure similarity.  
    - Checked for any skews, extreme observations, or discrepancies between the training and test sets that could impact model generalizability.

### Train Data
```{r}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(car)       # For assumption checks like residual plots
library(corrplot)  # For correlation matrix visualization
library(broom)     # For augmenting model output
library(dplyr)

# Define predictors and response variable
response <- "Quarterly_Average"
predictors <- setdiff(names(train_data), c("Quarter", response))

# Subset relevant data from the training set
eda_training_data <- train_data %>%
  select(all_of(c(response, predictors)))

@@

  print(p)  # Explicitly print the plot
}

### **2. Correlation Analysis in Training Set** ###
# Correlation matrix for the training set
cor_matrix_training <- cor(eda_training_data, use = "complete.obs")
print(cor_matrix_training)
corrplot(cor_matrix_training, method = "circle", type = "upper", tl.col = "black", tl.srt = 45)

### **3. Fit Linear Regression Model on Training Set** ###
# Fit the model using the training set
model_training <- lm(as.formula(paste(response, "~", paste(predictors, collapse = " + "))), data = eda_training_data)
summary(model_training)

# Extract residuals and fitted values from the training model
residuals_training <- resid(model_training)
fitted_values_training <- fitted(model_training)

# Calculate Variance Inflation Factors (VIF) for the training model
vif_training <- vif(model_training)
print(vif_training)

### **4. Outlier Detection in Training Set** ###
# Identify influential points using Cook's Distance
cooks_d_training <- cooks.distance(model_training)
influential_points_training <- which(cooks_d_training > 4 / (nrow(eda_training_data) - length(model_training$coefficients)))

# Output details of influential points
if (length(influential_points_training) > 0) {
  print(paste("Number of influential points:", length(influential_points_training)))
  print("Influential points identified:")
  print(influential_points_training)
} else {
  print("No influential points detected.")
}

```
### Test Data
```{r}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(car)       # For assumption checks like residual plots
library(corrplot)  # For correlation matrix visualization
library(broom)     # For augmenting model output
library(dplyr)

# Define predictors and response variable
response <- "Quarterly_Average"
predictors <- setdiff(names(test_data), c("Quarter", response))

# Subset relevant data from the test set
eda_test_data <- test_data %>%
  select(all_of(c(response, predictors)))

### Visualize Relationship
# Scatterplots for predictors in the test set
for (predictor in predictors) {
  p <- ggplot(eda_test_data, aes_string(x = predictor, y = response)) +
    geom_point(alpha = 0.7) +
    geom_smooth(method = "lm", color = "blue", se = FALSE) +
    labs(title = paste("Scatterplot of", response, "vs", predictor, "(Test Set)"),
         x = predictor, y = response) +
    theme_minimal()

  print(p)  # Explicitly print the plot
}

### Correlation Analysis
# Correlation matrix for the test set
cor_matrix_test <- cor(eda_test_data, use = "complete.obs")
print(cor_matrix_test)
corrplot(cor_matrix_test, method = "circle", type = "upper", tl.col = "black", tl.srt = 45)

### Linear Regression Model
# Fit the model using the test set
model_test <- lm(as.formula(paste(response, "~", paste(predictors, collapse = " + "))), data = eda_test_data)
summary(model_test)

# Extract residuals and fitted values from the test model
residuals_test <- resid(model_test)
fitted_values_test <- fitted(model_test)

# Calculate Variance Inflation Factors (VIF) for the test model
vif_test <- vif(model_test)
print(vif_test)

### Outlier Test Set
# Identify influential points using Cook's Distance
cooks_d_test <- cooks.distance(model_test)
influential_points_test <- which(cooks_d_test > 4 / (nrow(eda_test_data) - length(model_test$coefficients)))

# Output details of influential points
if (length(influential_points_test) > 0) {
  print(paste("Number of influential points:", length(influential_points_test)))
  print("Influential points identified:")
  print(influential_points_test)
} else {
  print("No influential points detected.")
}

@@

## **3. Model Building**
- **Tools Used**: `lm` in R for linear regression modeling.  
- **Steps Taken**:  
  - Conducted all model building, assessments, and diagnostics on the **train_data** to ensure the model generalizes well to unseen data.  
  - Included key predictors based on theoretical relevance and statistical significance.  
  - Applied transformations to predictors to improve linearity and stabilize variance.  
  - Addressed residual skewness with a Box-Cox transformation of the response variable.  
  - Selected the final model based on diagnostics from the training set, ensuring adherence to regression assumptions and a good fit.

```{r}
# Load necessary libraries
library(car)      # For VIF calculation
library(MASS)     # For Box-Cox transformation
library(lmtest)   # For diagnostic tests

@@

The multiple linear regression model for predicting the Box-Cox transformed quarterly average housing prices (\(bc\_Quarterly\_Average\)) shows a strong overall fit (\(R^2 = 0.9839\), \(Adj. R^2 = 0.9806\), \(F\)-statistic \(p < 2.2e-16\)), indicating that the selected predictors explain nearly all variability in the response variable. Key significant predictors include **Detached_Unabsorbed_Quarterly_Avg** (\(p = 8.58e-05\), positive effect), **Under_Construction_Detached** (\(p = 9.90e-06\), positive effect), and **CPI_Quarterly_Avg** (\(p = 2.34e-11\), strong positive effect). **Detached_Absorption_Quarterly_Avg** (\(p = 0.0293\)) exhibits a significant negative relationship with prices. While **Starting_Detached_Construction** (\(p = 0.0831\)) and **Completed_Construction_Semi** (\(p = 0.1387\)) are less significant, their inclusion is supported by theoretical relevance.

@@

Transformations were applied to stabilize variance, improve linearity, and address multicollinearity. Log-transformations were used for predictors like **Detached_Absorption_Quarterly_Avg**, **Starting_Detached_Construction**, and **Under_Construction_Detached** to handle skewness and make relationships with the response variable more linear. Standardizing **CPI_Quarterly_Avg** addressed extreme multicollinearity, as it was highly correlated with other predictors. A Box-Cox transformation was applied to the response variable to correct residual skewness and ensure homoscedasticity. These steps collectively improved model diagnostics, reduced variance inflation factors (VIFs), and ensured compliance with regression assumptions, resulting in a more robust and interpretable model.

## **4. Assessment of Assumptions**
- **Tools Used**: Residual diagnostics using residual vs. fitted plots, Q-Q plots, and histograms.  
- **Findings**:  
  - **Linearity**: Residual plots indicated clustering around fitted values, with minor deviations suggesting potential non-linearity.  
  - **Homoscedasticity**: Fanning in residuals for predictors
  - **Normality**: Q-Q plots looks for alignment with normality
  - **Influential Points**

```{r}
# Load necessary libraries
library(ggplot2)
library(car)
library(MASS)
library(lmtest)

# Calculate residuals and fitted values
residuals <- resid(model)
fitted_values <- fitted(model)

@@

## **5. Model Diagnostics**
- **Tools Used**: Performance metrics such as \( R^2 \), Adjusted \( R^2 \), AIC, BIC, and RMSE.  
- **Model Fit**:  
  - **\( R^2 \)**:
  - **Adjusted \( R^2 \)**
  - **AIC/BIC**:  
```{r}
# Load necessary libraries
library(Metrics)  # For RMSE calculation

# Model Performance Metrics
model_summary <- summary(model)

@@

## **6. Model Validation**

@@

# Step 1: Use the model to predict on test_data
predictions <- predict(model, newdata = test_data)

# Step 2: Calculate residuals
residuals <- test_data$Quarterly_Average - predictions
# Replace `Quarterly_Average` with the actual response variable name in test_data

# Step 3: Compute Mean Squared Error (MSE) for test_data
mse_test <- mean(residuals^2)
cat("Mean Squared Error (MSE) on Test Data:", mse_test, "\n")

# Step 4: Compute R-squared for test_data
ss_total <- sum((test_data$Quarterly_Average - mean(test_data$Quarterly_Average))^2)
ss_residual <- sum(residuals^2)
r_squared_test <- 1 - (ss_residual / ss_total)
cat("R-squared on Test Data:", r_squared_test, "\n")

# Step 7: Print validation results
cat("Validation Results:\n")
cat("Mean Squared Error (Test Data):", mse_test, "\n")
cat("R-squared (Test Data):", r_squared_test, "\n")

@@

## **7. Conclusion of Methods**

\newpage

>>>>

# Preliminary Results 
Focusing on the New Housing Price Index (NHPI) as the response variable, We will analyze economic, construction, and financial predictors such as construction activity, interest rates, and GDP.

## 1. Data Preparation

- Tools Used: Analysis was conducted in R using `tidyverse` for data manipulation and `car` for diagnostics.

- Steps Taken:  
  - Defined New Housing Price Index (NHPI) as the response variable and identified predictors:  
  1. Detached Absorption (Quarterly Average) 
  2. Semi Absorption (Quarterly Average)  
  3. Detached Unabsorbed Homes (Quarterly Average)  
  4. Semi Unabsorbed Homes (Quarterly Average)  
  5. GDP (Quarterly Average)  
  6. CPI (Quarterly Average) 
  7. Starting Detached Construction 
  8. Starting Semi-Detached Construction 
  9. Under Construction Detached Homes  
  10. Under Construction Semi-Detached Homes 
  11. Completed Construction Semi-Detached  
  12. Completed Construction Detached  
  13. Rate Hikes
  - Removed missing values and outliers using Cook’s Distance.
  - Merged datasets from 1997–2016 
  - Split the Data:  
    - Divided the dataset into training and test/validation sets.  
    - Used a common split, such as 50/50, to ensure both subsets are sufficiently large.  
    - Performed the split randomly to avoid sampling bias.  
```{r}
library(tidyverse)
library(here)
library(caret)
library(car)
library(corrplot)
library(broom)
library(MASS)
library(lmtest)
library(Metrics)

@@

## 2. Assumptions and Multicollinearity
- Tools Used: `ggplot2` for visualization 
- Key Insights:  
  - Scatterplots indicated potential linear relationships between predictors and NHPI.  
  - Variance Inflation Factors (VIFs) flagged multicollinearity among construction-related predictors, necessitating feature selection or transformation.  
  - Outliers were identified, influencing NHPI trends, and removed to enhance model accuracy.  
  - Explore Training and Test Sets:  
    - Performed EDA on both `training_set` and `test_set` to ensure consistency and reliability.  
    - Compared distributions, means, variances, and shapes of variables across both sets to ensure similarity.  
    - Checked for any skews, extreme observations, or discrepancies between the training and test sets that could impact model generalizability.

### Train Data
```{r}

# Predictors and response
response <- "Quarterly_Average"
predictors <- setdiff(names(train_data), c("Quarter", response))

# Data from the training set
eda_training_data <- train_data %>%
  select(all_of(c(response, predictors)))

@@

# Fit the model using the training set
model_training <- lm(as.formula(paste(response, "~", paste(predictors, collapse = " + "))), data = eda_training_data)
summary(model_training)

# Extract residuals and fitted values
residuals_training <- resid(model_training)
fitted_values_training <- fitted(model_training)

# Calculate VIF
vif_training <- vif(model_training)
print(vif_training)

# Influential points using Cook's Distance
cooks_d_training <- cooks.distance(model_training)
influential_points_training <- which(cooks_d_training > 4 / (nrow(eda_training_data) - length(model_training$coefficients)))

```
### Test Data
```{r}
response <- "Quarterly_Average"
predictors <- setdiff(names(test_data), c("Quarter", response))

# test set
eda_test_data <- test_data %>%
  select(all_of(c(response, predictors)))

# Scatterplots
for (predictor in predictors) {
  p <- ggplot(eda_test_data, aes_string(x = predictor, y = response)) +
    geom_point(alpha = 0.7) +
    geom_smooth(method = "lm", color = "blue", se = FALSE) +
    labs(title = paste("Scatterplot of", response, "vs", predictor, "(Test Set)"),
         x = predictor, y = response) +
    theme_minimal()

# Fit the model using the test set
model_test <- lm(as.formula(paste(response, "~", paste(predictors, collapse = " + "))), data = eda_test_data)
summary(model_test)

residuals_test <- resid(model_test)
fitted_values_test <- fitted(model_test)

# Calculate VIF 
vif_test <- vif(model_test)
print(vif_test)

# Identify influential points
cooks_d_test <- cooks.distance(model_test)
influential_points_test <- which(cooks_d_test > 4 / (nrow(eda_test_data) - length(model_test$coefficients)))

@@

## 3. Model Building
- Tools Used: `lm` in R for linear regression modeling.  
- Steps Taken:  
  - Conducted all model building, assessments, and diagnostics on the train_data to ensure the model generalizes well to unseen data.  
  - Included key predictors based on theoretical relevance and statistical significance.  
  - Applied transformations to predictors to improve linearity and stabilize variance.  
  - Addressed residual skewness with a Box-Cox transformation of the response variable.  
  - Selected the final model based on diagnostics from the training set, ensuring adherence to regression assumptions and a good fit.

```{r}
library(car)      
library(MASS)     
library(lmtest)

@@

The multiple linear regression model for predicting the Box-Cox transformed quarterly average housing prices (\(bc\_Quarterly\_Average\)) shows a strong overall fit (\(R^2 = 0.9839\), \(Adj. R^2 = 0.9806\), \(F\)-statistic \(p < 2.2e-16\)), indicating that the selected predictors explain nearly all variability in the response variable. Key significant predictors include Detached_Unabsorbed_Quarterly_Avg (\(p = 8.58e-05\), positive effect), Under_Construction_Detached (\(p = 9.90e-06\), positive effect), and CPI_Quarterly_Avg (\(p = 2.34e-11\), strong positive effect). Detached_Absorption_Quarterly_Avg (\(p = 0.0293\)) exhibits a significant negative relationship with prices. While Starting_Detached_Construction (\(p = 0.0831\)) and Completed_Construction_Semi (\(p = 0.1387\)) are less significant, their inclusion is supported by theoretical relevance.

@@

Transformations were applied to stabilize variance, improve linearity, and address multicollinearity. Log-transformations were used for predictors like Detached_Absorption_Quarterly_Avg, Starting_Detached_Construction, and Under_Construction_Detached to handle skewness and make relationships with the response variable more linear. Standardizing CPI_Quarterly_Avg addressed extreme multicollinearity, as it was highly correlated with other predictors. A Box-Cox transformation was applied to the response variable to correct residual skewness and ensure homoscedasticity. These steps collectively improved model diagnostics, reduced variance inflation factors (VIFs), and ensured compliance with regression assumptions, resulting in a more robust and interpretable model.

## 4. Assessment of Assumptions
- Tools Used: Residual diagnostics using residual vs. fitted plots, Q-Q plots, and histograms.  
- Findings:  
  - Linearity: Residual plots indicated clustering around fitted values, with minor deviations suggesting potential non-linearity.  
  - Homoscedasticity: Fanning in residuals for predictors
  - Normality: Q-Q plots looks for alignment with normality
  - Influential Points

```{r}
# Residuals and fitted values
residuals <- resid(model)
fitted_values <- fitted(model)

@@

## 5. Model Diagnostics
- Tools Used: Performance metrics such as \( R^2 \), Adjusted \( R^2 \), AIC, BIC, and RMSE.  
- Model Fit:  
  - \( R^2 \):
  - Adjusted \( R^2 \)
  - AIC/BIC:  
```{r}
# Model Performance Metrics
model_summary <- summary(model)

@@

## 6. Model Validation

@@

# Predict on test_data
predictions <- predict(model, newdata = test_data)

# Calculate residuals
residuals <- test_data$Quarterly_Average - predictions

# Compute MSE for test_data
mse_test <- mean(residuals^2)
cat("Mean Squared Error (MSE) on Test Data:", mse_test, "\n")

# R-squared for test_data
ss_total <- sum((test_data$Quarterly_Average - mean(test_data$Quarterly_Average))^2)
ss_residual <- sum(residuals^2)
r_squared_test <- 1 - (ss_residual / ss_total)
cat("R-squared on Test Data:", r_squared_test, "\n")

# Print validation results
cat("Validation Results:\n")
cat("Mean Squared Error (Test Data):", mse_test, "\n")
cat("R-squared (Test Data):", r_squared_test, "\n")

@@

## 7. Conclusion of Methods
The performance metrics demonstrate the effectiveness of the regression model on the test data. The Mean Squared Error (MSE) of 1.919877 reflects a small average error between the predicted and actual values, indicating high accuracy. The R-squared value of 0.993665 reveals that the model explains 99.37% of the variability in the test data, showcasing an excellent fit. These results confirm the model's strong predictive ability and validate its generalizability to unseen data.
\newpage
